
\chapter{Preface}

This \emph{Dhammapada Reflections}, Volume 2 follows on from Volume 1, which was printed in 2009. Once again it comprises a selection of 52 quotes from \emph{A Dhammapada for Contemplation} (2006) with comments. These quotes and comments were previously published as email messages sent out each new- and full-moon to participants in our Dhammasakaccha on-line correspondence community, which currently has about 600 participants. If you wish to receive these mailings please visit:

\vspace*{0.2\baselineskip}
{\centering
www.ratanagiri.org.uk/dhammapada-reflections/

}
\vspace*{0.2\baselineskip}

Once again, I am indebted to many who have helped in the preparation of this material. For the Dhammapada verses themselves I consulted several authoritative versions. In particular I have used the work by Ven. Narada Thera (B.M.S. 1978), Ven. Ananda Maitreya Thera (Lotsawa 1988), Daw Mya Tin and the editors of the Burmese Pitaka Association (1987) and Ajahn Thanissaro. For the recorded stories associated with the verses I also turned to www.tipitaka.net. 

Ron Lumsden and Jayasiri Lewis have generously given of their time and skill in editing the texts.

May the blessings arising from this compilation be shared with all who have been involved in its production and sponsorship.  May all who seek the way find it and experience the freedom at its end. May all beings seek the way.

{\par\raggedleft
Bhikkhu Munindo\\
Aruna Ratanagiri Buddhist Monastery\\
Northumberland, UK
\par}

